---
title: "Tr1: Enrichment Analysis"
author: "Stephen Pederson & Caitlin Abbott"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 8,
  dev = c("png", "pdf")
)
```


```{r packages}
library(tidyverse)
library(limma)
library(edgeR)
library(RColorBrewer)
library(pheatmap)
library(magrittr)
library(clusterProfiler)
library(org.Mm.eg.db)
library(reactable)
library(htmltools)
library(glue)
library(scales)
with_tooltip <- function(value, width = 30) {
  tags$span(title = value, str_trunc(value, width))
}
```

```{r load-data}
top_tables <- read_rds(here::here("output", "top_tables.rds"))
dgeFilt <- read_rds(here::here("output", "dgeFilt.rds"))
```


```{r universe}
universe <- dgeFilt$genes$entrezid %>%
  unlist() %>%
  extract(!is.na(.)) %>% #keep non NA gene
  unique() %>%
  as.character()
```

## Enrichment Results {.tabset}

```{r ego-results}
ego_results <- top_tables %>% 
  lapply(
    function(x) {
      de <- dplyr::filter(x, adj.P.Val < 0.05, abs(logFC) > 1)
      ids <- unique(unlist(de$entrezid))
      enrichGO(
        ids,
        universe = universe,
        OrgDb = org.Mm.eg.db,
        ont = "ALL",
        keyType = "ENTREZID",
        pAdjustMethod = "bonferroni",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.05,
        readable = TRUE
      )
    }
  )
```

Enrichment testing was performed using all three available Gene Ontologies and the function `enrichGO()` from the package `clusterProfiler`.
Ontologies were only considered to be enriched amongst the differentially expressed genes if a Bonferroni-adjusted p-value < 0.05 was returned during enrichment testing, with no regard to the sign of fold change.

### DN Vs. Lag3^+^

```{r dotplot-dn-lag3}
dotplot(ego_results$DNvLAG3) +
  ggtitle(expression(paste("DN Vs. Lag", 3^{textstyle("+")}))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### DN Vs. DP

```{r dotplot-dn-dp}
dotplot(ego_results$DNvDP) +
  ggtitle("DN Vs. DP") +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### DN Vs. Cd49b^+^

```{r dotplot-dn-cd49}
dotplot(ego_results$DNvCD49b) +
  ggtitle(expression(paste("DN Vs. Cd49", b^{textstyle("+")}))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### Lag3^+^ Vs Cd49^+^

```{r dotplot-lag3-cd49}
dotplot(ego_results$LAG3vCD49b) +
  ggtitle(
    expression(
      paste("Lag", 3^{textstyle("+")}, " Vs. CD49", b^{textstyle("+")})
    )
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### Lag3^+^ Vs. DP

```{r dotplot-lag3-dp}
dotplot(ego_results$LAG3vDP) +
  ggtitle(expression(paste("Lag", 3^{textstyle("+")}, " Vs. DP"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```


### Cd49b^+^ Vs. DP

```{r dotplot-cd49-dp}
dotplot(ego_results$CD49bvDP) +
  ggtitle(expression(paste("Cd49", b^{textstyle("+")}, " Vs. DP"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

## Enrichment Tables {.tabset}

```{r enrichment-tables}
htmltools::tagList(
  names(ego_results) %>% 
    lapply(
      function(x) {
        df <- ego_results[[x]] %>% 
          as_tibble() %>% 
          dplyr::filter(p.adjust < 0.05) %>% 
          mutate(
            geneID = str_replace_all(geneID, "\\/", "; "),
            `% DE` = vapply(GeneRatio, function(x) eval(parse(text = x)), numeric(1)),
            `% BG` = vapply(BgRatio, function(x) eval(parse(text = x)), numeric(1))
          ) %>% 
          dplyr::select(ONTOLOGY, ID, Description, Count, starts_with("%"), p.adjust, geneID) %>% 
          arrange(p.adjust) %>% 
          distinct(ONTOLOGY, geneID, .keep_all = TRUE) 
        
        if (nrow(df)) {
          cp <- glue(
            "All {nrow(df)} ontologies considered enriched in the set of ",
            length(unique(unlist(dplyr::filter(top_tables[[x]], DE)$entrezid))),
            " DE genes mapped to EntrezIDs from the comparison {x}. Enrichment ",
            "testing was performed using Fisher's Exact Test compared to the set of ",
            comma(length(unique(universe))), 
            " unique EntrezIDs mapped to genes considered as detectable. ",
            "The percentage of nonDE genes mapped to the ontology are also given. ",
            "P-values are Bonferroni adjusted and all pathways are considered enriched ",
            "using an adjusted p-value < 0.05. In order to see all DE genes, hover your ", 
            "mouse over the final column. ",
            "Where the identical genes map to multiple terms within each ontology, only ", 
            "the term with the lowest p-value (i.e. strongest enrichment) is shown. ",
            "As a result, some of the pathways seen in the above dotplots may not be ",
            "presented in these tables."
          )
          htmltools::div(
            htmltools::div(
              id = x, 
              class="section level3",
              htmltools::h3(class = "tabset", x),
              htmltools::tags$em(cp),
              df %>% 
                reactable(
                  filterable = TRUE,
                  columns = list(
                    ONTOLOGY = colDef(name = "Ontology", maxWidth = 100),
                    ID = colDef(name = "GO ID", maxWidth = 125),
                    Description = colDef(maxWidth = 250),
                    Count = colDef(name = "N<sub>DE</sub>", html = TRUE, maxWidth = 80),
                    `% DE` = colDef(format = colFormat(percent = TRUE, digits = 1), maxWidth = 85),
                    `% BG` = colDef(format = colFormat(percent = TRUE, digits = 1), maxWidth = 85),
                    p.adjust = colDef(
                      name = "p<sub>adj</sub>", html = TRUE, maxWidth = 100,
                      cell = function(value) {
                        fm <- ifelse(value < 0.001, "%.2e", "%.3f")
                        sprintf(fm, value)
                      }
                    ),
                    geneID = colDef(
                      name = "DE Genes",
                      cell = function(value) with_tooltip(value, width = 100)
                    )
                  )
                )
            )
          )
        } else {
          cat(glue("No enrichment was found for {x}\n\n"))
        }
      }
    )
)
```


## Enrichment Results (Up Only) {.tabset}

```{r ego-results-up}
ego_results <- top_tables %>% 
  lapply(
    function(x) {
      de <- dplyr::filter(x, adj.P.Val < 0.05, logFC > 1)
      ids <- unique(unlist(de$entrezid))
      enrichGO(
        ids,
        universe = universe,
        OrgDb = org.Mm.eg.db,
        ont = "ALL",
        keyType = "ENTREZID",
        pAdjustMethod = "bonferroni",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.05,
        readable = TRUE
      )
    }
  )
```

Enrichment testing was repeated using all three available Gene Ontologies and the function `enrichGO()` from the package `clusterProfiler`.
Ontologies were only considered to be enriched amongst the **up-regulated genes** if a Bonferroni-adjusted p-value < 0.05 was returned during enrichment testing.
Please note that an up-regulated gene does not always correspond to an increase in activity of a pathway.

### DN Vs. Lag3^+^

```{r dotplot-dn-lag3-up}
dotplot(ego_results$DNvLAG3) +
  ggtitle(expression(paste("DN Vs. Lag", 3^{textstyle("+")}, "(Up Only)"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### DN Vs. DP

*No significant results were obtained for up-regulated genes in this comparison.*

```{r dotplot-dn-dp-up, eval=FALSE}
dotplot(ego_results$DNvDP) +
  ggtitle("DN Vs. DP (Up Only)") +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### DN Vs. Cd49b^+^

```{r dotplot-dn-cd49-up}
dotplot(ego_results$DNvCD49b) +
  ggtitle(expression(paste("DN Vs. Cd49", b^{textstyle("+")}, "(Up Only)"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### Lag3^+^ Vs Cd49^+^

```{r dotplot-lag3-cd49-up}
dotplot(ego_results$LAG3vCD49b) +
  ggtitle(
    expression(
      paste("Lag", 3^{textstyle("+")}, " Vs. CD49", b^{textstyle("+")}, "(Up Only)")
    )
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```


### Lag3^+^ Vs. DP

*No significant results were obtained for up-regulated genes in this comparison.*

```{r dotplot-lag3-dp-up, eval = FALSE}
dotplot(ego_results$LAG3vDP) +
  ggtitle(expression(paste("Lag", 3^{textstyle("+")}, " Vs. DP", "(Up Only)"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```


### Cd49b^+^ Vs. DP

```{r dotplot-cd49-dp-up}
dotplot(ego_results$CD49bvDP) +
  ggtitle(expression(paste("Cd49", b^{textstyle("+")}, " Vs. DP (Up Only)"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```


## Enrichment Tables (Up Only) {.tabset}

```{r enrichment-tables-up}
ego_results <- ego_results[vapply(ego_results, nrow, integer(1)) > 0]
htmltools::tagList(
  names(ego_results) %>% 
    lapply(
      function(x) {
        df <- ego_results[[x]] %>% 
          as_tibble() %>% 
          dplyr::filter(p.adjust < 0.05) %>% 
          mutate(
            geneID = str_replace_all(geneID, "\\/", "; "),
            `% DE` = vapply(GeneRatio, function(x) eval(parse(text = x)), numeric(1)),
            `% BG` = vapply(BgRatio, function(x) eval(parse(text = x)), numeric(1))
          ) %>% 
          dplyr::select(ONTOLOGY, ID, Description, Count, starts_with("%"), p.adjust, geneID) %>% 
          arrange(p.adjust) %>% 
          distinct(ONTOLOGY, geneID, .keep_all = TRUE) 
        
        if (nrow(df)) {
          cp <- glue(
            "All {nrow(df)} ontologies considered enriched in the set of ",
            length(unique(unlist(dplyr::filter(top_tables[[x]], DE, logFC > 1)$entrezid))),
            " up-regulated genes mapped to EntrezIDs from the comparison {x}. Enrichment ",
            "testing was performed using Fisher's Exact Test compared to the set of ",
            comma(length(unique(universe))), 
            " unique EntrezIDs mapped to genes considered as detectable. ",
            "The percentage of nonDE genes mapped to the ontology are also given. ",
            "P-values are Bonferroni adjusted and all pathways are considered enriched ",
            "using an adjusted p-value < 0.05. In order to see all DE genes, hover your ", 
            "mouse over the final column. ",
            "Where the identical genes map to multiple terms within each ontology, only ", 
            "the term with the lowest p-value (i.e. strongest enrichment) is shown. ",
            "As a result, some of the pathways seen in the above dotplots may not be ",
            "presented in these tables."
          )
          htmltools::div(
            htmltools::div(
              id = x, 
              class="section level3",
              htmltools::h3(class = "tabset", x),
              htmltools::tags$em(cp),
              df %>% 
                reactable(
                  filterable = TRUE,
                  columns = list(
                    ONTOLOGY = colDef(name = "Ontology", maxWidth = 100),
                    ID = colDef(name = "GO ID", maxWidth = 125),
                    Description = colDef(maxWidth = 250),
                    Count = colDef(name = "N<sub>DE</sub>", html = TRUE, maxWidth = 80),
                    `% DE` = colDef(format = colFormat(percent = TRUE, digits = 1), maxWidth = 85),
                    `% BG` = colDef(format = colFormat(percent = TRUE, digits = 1), maxWidth = 85),
                    p.adjust = colDef(
                      name = "p<sub>adj</sub>", html = TRUE, maxWidth = 100,
                      cell = function(value) {
                        fm <- ifelse(value < 0.001, "%.2e", "%.3f")
                        sprintf(fm, value)
                      }
                    ),
                    geneID = colDef(
                      name = "DE Genes",
                      cell = function(value) with_tooltip(value, width = 100)
                    )
                  )
                )
            )
          )
        } else {
          cat(glue("No enrichment was found for {x}\n\n"))
        }
      }
    )
)
```


## Enrichment Results (Down Only) {.tabset}

```{r ego-results-down}
ego_results <- top_tables %>% 
  lapply(
    function(x) {
      de <- dplyr::filter(x, adj.P.Val < 0.05, logFC < -1)
      ids <- unique(unlist(de$entrezid))
      enrichGO(
        ids,
        universe = universe,
        OrgDb = org.Mm.eg.db,
        ont = "ALL",
        keyType = "ENTREZID",
        pAdjustMethod = "bonferroni",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.05,
        readable = TRUE
      )
    }
  )
```

Enrichment testing was repeated using all three available Gene Ontologies and the function `enrichGO()` from the package `clusterProfiler`.
Ontologies were only considered to be enriched amongst the **down-regulated genes** if a Bonferroni-adjusted p-value < 0.05 was returned during enrichment testing.
Please note that a down-regulated gene does not always correspond to an decrease in activity of a pathway.

### DN Vs. Lag3^+^

```{r dotplot-dn-lag3-down}
dotplot(ego_results$DNvLAG3) +
  ggtitle(expression(paste("DN Vs. Lag", 3^{textstyle("+")}, "(Down Only)"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### DN Vs. DP

```{r dotplot-dn-dp-down}
dotplot(ego_results$DNvDP) +
  ggtitle("DN Vs. DP (Down Only)") +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### DN Vs. Cd49b^+^

```{r dotplot-dn-cd49-down}
dotplot(ego_results$DNvCD49b) +
  ggtitle(expression(paste("DN Vs. Cd49", b^{textstyle("+")}, "(Down Only)"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### Lag3^+^ Vs Cd49^+^

```{r dotplot-lag3-cd49-down}
dotplot(ego_results$LAG3vCD49b) +
  ggtitle(
    expression(
      paste("Lag", 3^{textstyle("+")}, " Vs. CD49", b^{textstyle("+")}, "(Down Only)")
    )
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```


### Lag3^+^ Vs. DP

```{r dotplot-lag3-dp-down}
dotplot(ego_results$LAG3vDP) +
  ggtitle(expression(paste("Lag", 3^{textstyle("+")}, " Vs. DP", "(Down Only)"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```


### Cd49b^+^ Vs. DP

```{r dotplot-cd49-dp-down}
dotplot(ego_results$CD49bvDP) +
  ggtitle(expression(paste("Cd49", b^{textstyle("+")}, " Vs. DP (Down Only)"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

## Enrichment Tables (Down Only) {.tabset}

```{r enrichment-tables-down}
ego_results <- ego_results[vapply(ego_results, nrow, integer(1)) > 0]
htmltools::tagList(
  names(ego_results) %>% 
    lapply(
      function(x) {
        df <- ego_results[[x]] %>% 
          as_tibble() %>% 
          dplyr::filter(p.adjust < 0.05) %>% 
          mutate(
            geneID = str_replace_all(geneID, "\\/", "; "),
            `% DE` = vapply(GeneRatio, function(x) eval(parse(text = x)), numeric(1)),
            `% BG` = vapply(BgRatio, function(x) eval(parse(text = x)), numeric(1))
          ) %>% 
          dplyr::select(ONTOLOGY, ID, Description, Count, starts_with("%"), p.adjust, geneID) %>% 
          arrange(p.adjust) %>% 
          distinct(ONTOLOGY, geneID, .keep_all = TRUE) 
        
        if (nrow(df)) {
          cp <- glue(
            "All {nrow(df)} ontologies considered enriched in the set of ",
            length(unique(unlist(dplyr::filter(top_tables[[x]], DE, logFC < -1)$entrezid))),
            " down-regulated genes mapped to EntrezIDs from the comparison {x}. Enrichment ",
            "testing was performed using Fisher's Exact Test compared to the set of ",
            comma(length(unique(universe))), 
            " unique EntrezIDs mapped to genes considered as detectable. ",
            "The percentage of nonDE genes mapped to the ontology are also given. ",
            "P-values are Bonferroni adjusted and all pathways are considered enriched ",
            "using an adjusted p-value < 0.05. In order to see all DE genes, hover your ", 
            "mouse over the final column. ",
            "Where the identical genes map to multiple terms within each ontology, only ", 
            "the term with the lowest p-value (i.e. strongest enrichment) is shown. ",
            "As a result, some of the pathways seen in the above dotplots may not be ",
            "presented in these tables."
          )
          htmltools::div(
            htmltools::div(
              id = x, 
              class="section level3",
              htmltools::h3(class = "tabset", x),
              htmltools::tags$em(cp),
              df %>% 
                reactable(
                  filterable = TRUE,
                  columns = list(
                    ONTOLOGY = colDef(name = "Ontology", maxWidth = 100),
                    ID = colDef(name = "GO ID", maxWidth = 125),
                    Description = colDef(maxWidth = 250),
                    Count = colDef(name = "N<sub>DE</sub>", html = TRUE, maxWidth = 80),
                    `% DE` = colDef(format = colFormat(percent = TRUE, digits = 1), maxWidth = 85),
                    `% BG` = colDef(format = colFormat(percent = TRUE, digits = 1), maxWidth = 85),
                    p.adjust = colDef(
                      name = "p<sub>adj</sub>", html = TRUE, maxWidth = 100,
                      cell = function(value) {
                        fm <- ifelse(value < 0.001, "%.2e", "%.3f")
                        sprintf(fm, value)
                      }
                    ),
                    geneID = colDef(
                      name = "DE Genes",
                      cell = function(value) with_tooltip(value, width = 100)
                    )
                  )
                )
            )
          )
        } else {
          cat(glue("No enrichment was found for {x}\n\n"))
        }
      }
    )
)
```
