---
title: "Tr1: Enrichment Analysis"
author: "Stephen Pederson & Caitlin Abbott"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 8,
  dev = c("png", "pdf")
)
```


```{r packages}
library(tidyverse)
library(limma)
library(edgeR)
library(RColorBrewer)
library(pheatmap)
library(magrittr)
library(clusterProfiler)
library(org.Mm.eg.db)
```

```{r load-data}
top_tables <- read_rds(here::here("output", "top_tables.rds"))
dgeFilt <- read_rds(here::here("output", "dgeFilt.rds"))
```


```{r universe}
universe <- dgeFilt$genes$entrezid %>%
  unlist() %>%
  extract(!is.na(.)) %>% #keep non NA gene
  unique() %>%
  as.character()
```

```{r ego-results}
ego_results <- top_tables %>% 
  lapply(
    function(x) {
      de <- dplyr::filter(x, adj.P.Val < 0.05, abs(logFC) > 1)
      ids <- unique(unlist(de$entrezid))
      enrichGO(
        ids,
        universe = universe,
        OrgDb = org.Mm.eg.db,
        ont = "ALL",
        keyType = "ENTREZID",
        pAdjustMethod = "bonferroni",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.05,
        readable = TRUE
      )
    }
  )
```

## Enrichment Results {.tabset}

Enrichment testing was performed using all three available Gene Ontologies and the function `enrichGO()` from the package `clusterProfiler`.
Ontologies were only considered to be enriched amongst the differentially expressed genes if a Bonferroni-adjusted p-value < 0.05 was returned during enrichment testing.

### DN Vs. Lag3^+^

```{r dotplot-dn-lag3}
dotplot(ego_results$DNvLAG3) +
  ggtitle(expression(paste("DN Vs. Lag", 3^{textstyle("+")}))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### DN Vs. DP

```{r dotplot-dn-dp}
dotplot(ego_results$DNvDP) +
  ggtitle("DN Vs. DP") +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### DN Vs. Cd49b^+^

```{r dotplot-dn-cd49}
dotplot(ego_results$DNvCD49b) +
  ggtitle(expression(paste("DN Vs. Cd49", b^{textstyle("+")}))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```

### Lag3^+^ Vs Cd49^+^

```{r dotplot-lag3-cd49}
dotplot(ego_results$LAG3vCD49b) +
  ggtitle(
    expression(
      paste("Lag", 3^{textstyle("+")}, " Vs. CD49", b^{textstyle("+")})
    )
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```


### Lag3^+^ Vs. DP

```{r dotplot-lag3-dp}
dotplot(ego_results$LAG3vDP) +
  ggtitle(expression(paste("Lag", 3^{textstyle("+")}, " Vs. DP"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```


### Cd49b^+^ Vs. DP

```{r dotplot-cd49-dp}
dotplot(ego_results$CD49bvDP) +
  ggtitle(expression(paste("Cd49", b^{textstyle("+")}, " Vs. DP"))) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 16)
  )
```
